Pillar 1 - Sequential Pattern Mining

### 2019-02-11 Pillar 1 - Sequential Pattern Mining.Rmd
use of lists to simplify environment

@@

### take, fail, low grade
```{r AR data prep 1}

#
# Threshold: pass grade, high grade
pass_grade <- 5.5
high_grade <- 6.5

#
# Transactions
d_transactions <- d_course %>%

@@

### take / not take, pass / fail / not take
```{r transactions not taken}
d_transactions_TPF <- expand.grid(

  # Expand along students (sequenceID) and courses (itemID)
  sequenceID = unique(d_transactions$sequenceID),
  item       = unique(d_transactions$item),
  stringsAsFactors = FALSE
  ) %>%

  # Join with d_transactions
  left_join(
    d_transactions, 
    by = c("sequenceID", "item")
    ) %>%

@@

d_transactions_G <- d_transactions %>%

@@

#
# Number of students
n_students <- d_transactions %>%
  select(sequenceID) %>%
  n_distinct

#
# Probability of taking a course
d_support_TNT <- d_transactions %>%
  distinct(
    item,
    sequenceID
    ) %>%
  count(
    item
    ) %>%
  mutate(
    rate.take = n / n_students,
    rate.not  = 1 - rate.take
    )

#
# Probability of failing, having a low grade
d_support_PF_HL <- d_transactions %>%
  group_by(
    item
    ) %>%
  summarise(
    rate.fail = mean(PF == "fail"),
    rate.low  = mean(HL == "low")
    )

#
# Probability oh having a grade equal or lower than x
d_support_G <- d_transactions %>%
  group_by(
    item
    ) %>%
  summarise(
    `0`    = mean(Grade <= 0),
    `1`    = mean(Grade <= 1),
    `2`    = mean(Grade <= 2),
    `3`    = mean(Grade <= 3),
    `4`    = mean(Grade <= 4),
    `5`    = mean(Grade <= 5),
    `6`    = mean(Grade <= 6),
    `7`    = mean(Grade <= 7),
    `8`    = mean(Grade <= 8),
    `9`    = mean(Grade <= 9),
    `10`   = mean(Grade <= 10)
    ) %>%
  gather(
    key   = grade_ceil,
    value = support.grade,
    `0` : `10`,
    convert = TRUE
    )

@@

make_transaction <- function(data = d_transactions, item = item){

  data %>%
    group_by(
      sequenceID
    ) %>%
    summarise(
      list_item = list(!!enquo(item))
    ) %>%
    ungroup %>%
    pull(
      list_item
      ) %>%
    as("transactions")

@@

#
# Making transactions
transactions     <- make_transaction(item = item   )
transactions_PF  <- make_transaction(item = item_PF)
transactions_HL  <- make_transaction(item = item_HL)
transactions_TPF <- make_transaction(data = d_transactions_TPF, item = item_TPF)
transactions_G   <- make_transaction(data = d_transactions_G  , item = item_G  )

@@

make_sequence <- function(data = d_transactions, item = item){

@@

#
# Making sequences
sequences     <- make_sequence(item = item   )
sequences_PF  <- make_sequence(item = item_PF)
sequences_HL  <- make_sequence(item = item_HL)
#sequences_TPF <- make_sequence(data = d_transactions_TPF, item = item_TPF) # compute ad hoc
sequences_G   <- make_sequence(data = d_transactions_G    , item = item_G  )

@@

rm(make_sequence,
   d_transactions, d_transactions_TPF, d_transactions_G)
```

@@

```{r editing_AR}
editing_AR <- function(AR){

  AR %>%
    mutate_at(
      c("support", 
        "confidence", "rhs.support",
        "lift"
        ),
      funs(round(., 5))
    ) %>%
    select(
      lhs, rhs,
      support, count,
      confidence, rhs.support,
      lift
    ) %>%
    arrange(
      desc(count)
    )

@@

We encapsulate the four functions `my_apriori()`, `clean_AR()`, `compute_support()` and `editing_AR()` into the function `make_AR()`.

  data %>%
    my_apriori %>%
    clean_AR %>%
    compute_metrics(
      data_support = data_support,
      type_rule = !!enquo(type_rule)
      ) %>%
    editing_AR

```{r AR take}
AR_taken <- transactions %>%
  make_AR(
    data_support = d_support_TNT, 
    type_rule = rate.take
    )

rm(transactions, d_support_TNT)
```

```{r AR fail}
AR_PF <- transactions_PF %>%

  make_AR(
    data_support = d_support_PF_HL,
    type_rule = rate.fail
    ) %>%

@@

  # lhs and rhs must be fail, count >= 5
  filter(
    str_detect(lhs, "fail"),
    str_detect(rhs, "fail"),
    count >= 5
    ) %>%
  select(
    - c(rhs_outcome, rhs_course)
  )

rm(transactions_PF)
```

```{r AR low}
AR_HL <- transactions_HL %>%

  make_AR(
    data_support = d_support_PF_HL,
    type_rule = rate.low
    ) %>%

@@

  # lhs and rhs must be fail, count >= 5
  filter(
    str_detect(lhs, "low"),
    str_detect(rhs, "low"),
    count >= 5
    ) %>%
  select(
    - c(rhs_outcome, rhs_course)
  )

rm(transactions_HL)
```

```{r AR TPF}
AR_TPF <- transactions_TPF %>%

  make_AR(
    data_support = d_support_PF_HL, 
    type_rule = rate.fail
  ) %>%

  # Confidence and Lift
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  mutate(
    rhs.take = str_detect(rhs, "fail|pass")
    ) %>%
  group_by(
    lhs, rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = sum(support * rhs.take),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

  # lhs and rhs must be fail, count >= 5
  filter(
    str_detect(lhs, "not taken"),
    str_detect(rhs, "fail"),
    count >= 5
    ) %>%
  select(
    - c(rhs_outcome, rhs_course, rhs.take)
  )

rm(transactions_TPF, d_support_PF_HL)
```

```{r AR grade}
AR_G <- transactions_G %>%

@@

  # rhs.support
  left_join(
    d_support_G,
    by = c("rhs_course" = "item", "rhs_grade" = "grade_ceil")
    ) %>%
  mutate(
    rhs.support = support.grade
    ) %>%

  # Confidence and lift
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  group_by(
    lhs, 
    rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = max(support),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

  # reduce number of rules
  filter(
    lhs_course != rhs_course, # exclude rules with same course on lhs and rhs
    count >= 5,               
    rhs_grade == 6            
    ) %>%

  # for each combination of courses (lhs and rhs), keep AR with highest lift (most informative AR)
  group_by(
    lhs_course, 
    rhs_course
    ) %>%
  top_n(
    1,
    lift
    ) %>%
  ungroup %>%

  editing_AR

rm(transactions_G, d_support_G)
```

We save and remove objects:
```{r saving and rm}
#
# Save association rules
save(AR_taken, AR_PF, AR_HL, AR_TPF, AR_G,
     file = "App/AR.RDATA")

# Remove objects
rm(my_apriori, clean_AR, compute_metrics, editing_AR, make_AR,
   AR_taken, AR_PF, AR_HL, AR_TPF, AR_G,
   n_students)
```

>>>>

```{r}
pass_grade <- 5.5
high_grade <- 6.5
```

@@

### take, fail, low grade
```{r AR data prep 1}
d_transactions <- list()

d_transactions$taken_PF_HL <- d_course %>%

@@

### take / not take, pass / fail / not take
```{r transactions not taken}
d_transactions$TPF <- expand.grid(

  # Expand along students (sequenceID) and courses (itemID)
  sequenceID = unique(d_transactions$taken_PF_HL$sequenceID),
  item       = unique(d_transactions$taken_PF_HL$item),
  stringsAsFactors = FALSE
  ) %>%

  # Join with d_transactions
  left_join(
    d_transactions$taken_PF_HL, 
    by = c("sequenceID", "item")
    ) %>%

@@

d_transactions$G <- d_transactions$taken_PF_HL %>%

@@

#
# Number of students
n_students <- d_transactions$taken_PF_HL %>%
  select(sequenceID) %>%
  n_distinct

#
# Support
d_support <- list()

# Probability of taking a course
d_support$TNT <- d_transactions$taken_PF_HL %>%
  distinct(
    item,
    sequenceID
    ) %>%
  count(
    item
    ) %>%
  mutate(
    rate.take = n / n_students,
    rate.not  = 1 - rate.take
    )

#
# Probability of failing, having a low grade
d_support$PF_HL <- d_transactions$taken_PF_HL %>%
  group_by(
    item
    ) %>%
  summarise(
    rate.fail = mean(PF == "fail"),
    rate.low  = mean(HL == "low")
    )

#
# Probability oh having a grade equal or lower than x
d_support$G <- d_transactions$taken_PF_HL %>%
  group_by(
    item
    ) %>%
  summarise(
    `0`    = mean(Grade <= 0),
    `1`    = mean(Grade <= 1),
    `2`    = mean(Grade <= 2),
    `3`    = mean(Grade <= 3),
    `4`    = mean(Grade <= 4),
    `5`    = mean(Grade <= 5),
    `6`    = mean(Grade <= 6),
    `7`    = mean(Grade <= 7),
    `8`    = mean(Grade <= 8),
    `9`    = mean(Grade <= 9),
    `10`   = mean(Grade <= 10)
    ) %>%
  gather(
    key   = grade_ceil,
    value = support.grade,
    `0` : `10`,
    convert = TRUE
    )

@@

make_transaction <- function(data = d_transactions$taken_PF_HL, item = item){

  data %>%
    group_by(
      sequenceID
    ) %>%
    summarise(
      list_item = list(unique(!!enquo(item)))
    ) %>%
    ungroup %>%
    pull(
      list_item
      ) %>%
    as("transactions")

@@

#
# Making transactions
transactions <- list()

transactions$taken <- make_transaction(item = item   )
transactions$PF    <- make_transaction(item = item_PF)
transactions$HL    <- make_transaction(item = item_HL)
transactions$TPF   <- make_transaction(data = d_transactions$TPF, item = item_TPF)
transactions$G     <- make_transaction(data = d_transactions$G  , item = item_G  )

@@

make_sequence <- function(data = d_transactions$taken_PF_HL, item = item){

@@

#
# Making sequences
sequences <- list()

sequences$taken <- make_sequence(item = item   )
sequences$PF    <- make_sequence(item = item_PF)
sequences$HL    <- make_sequence(item = item_HL)
#sequences_TPF <- make_sequence(data = d_transactions_TPF, item = item_TPF) # compute ad hoc
sequences$G   <- make_sequence(data = d_transactions$G    , item = item_G  )

@@

rm(make_sequence, d_transactions)
```

@@

We encapsulate the functions `my_apriori()`, `clean_AR()` and `compute_support()` into the function `make_AR()`.

  data %>%
    my_apriori %>%
    clean_AR %>%
    compute_metrics(
      data_support = data_support,
      type_rule = !!enquo(type_rule)
      )

@@

```{r AR take}
AR <- list()

AR$taken <- transactions$taken %>%
  make_AR(
    data_support = d_support$TNT, 
    type_rule = rate.take
    )

```{r AR fail}
AR$PF <- transactions$PF %>%

  make_AR(
    data_support = d_support$PF_HL,
    type_rule = rate.fail
    ) %>%

@@

  # lhs and rhs must be fail, count >= 5
  filter(
    str_detect(lhs, "fail"),
    str_detect(rhs, "fail")
    )
```

```{r AR low}
AR$HL <- transactions$HL %>%

  make_AR(
    data_support = d_support$PF_HL,
    type_rule = rate.low
    ) %>%

@@

  # lhs and rhs must be fail, count >= 5
  filter(
    str_detect(lhs, "low"),
    str_detect(rhs, "low")
    )
```

```{r AR TPF}
AR$TPF <- transactions$TPF %>%

  make_AR(
    data_support = d_support$PF_HL, 
    type_rule = rate.fail
  ) %>%

  # Confidence and Lift
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  mutate(
    rhs.taken = str_detect(rhs, "fail|pass")
    ) %>%
  group_by(
    lhs, rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = sum(support * rhs.taken),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

  # lhs and rhs must be fail, count >= 5
  filter(
    str_detect(lhs, "not taken"),
    str_detect(rhs, "fail")
    )
```

```{r AR grade}
AR$G <- transactions$G %>%

@@

  # rhs.support
  left_join(
    d_support$G,
    by = c("rhs_course" = "item", "rhs_grade" = "grade_ceil")
    ) %>%
  mutate(
    rhs.support = support.grade
    ) %>%

  # Confidence and lift
  group_by(
    lhs, 
    rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = max(support),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

  # reduce number of rules
  filter(
    lhs_course != rhs_course, # exclude rules with same course on lhs and rhs        
    rhs_grade == 6            
    ) %>%

  # for each combination of courses (lhs and rhs), keep AR with highest lift (most informative AR)
  group_by(
    lhs_course, 
    rhs_course
    ) %>%
  top_n(
    1,
    lift
    ) %>%
  ungroup
```

```{r edit_AR}
edit_AR <- function(AR){

  AR %>%
    filter(
      count >= 5
      ) %>%
    mutate_if(
      is.numeric,
      funs(round(., 5))
    ) %>%
    select(
      lhs, rhs,
      support, count,
      confidence, rhs.support,
      lift, 
      everything(),
      - matches("rate|course|grade|outcome|taken")
    ) %>%
    arrange(
      desc(count)
    )

AR <- lapply(
  X   = AR,
  FUN = edit_AR
    )
```

We save and remove objects:
```{r saving and rm}
#
# Save association rules
save(AR,
     file = "App/AR.RDATA")

# Remove objects
rm(my_apriori, clean_AR, compute_metrics, edit_AR, make_AR,
   AR)
```
